---
title: "A simple tutorial using SpatialQPFs library to extract spatial features based on cell level data from images"
author: "Xiao Li <xiao.li.xl2@roche.com>"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A simple tutorial using SpatialQPFs library}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, include = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, dpi = 100, fig.width = 5, fig.height = 5)
```


# Introduction
This Rmarkdown file provides a simple tutorial on how to use the functions in SpatialQPFs R library to calculate the features to decipher spatial relationship between different cell types, from spatial statistics and graph topology perspective.  





# How to use the functions in SpatialQPFs
Firstly, let's simulate a synthetic dataset to begin with. In reality, to extract detailed cell-level information, users must preprocess the digital image. This preprocessing step involves isolating key information such as cell coordinates and cell identity from a variety of digital slides, which can encompass images stained with H$\&$E, Immunohistochemistry, or Immunofluorescence techniques, as well as spatially resolved single-cell imaging. 



```{r, warning = FALSE}
library('SpatialQPFs')
library('data.table')
library('dplyr')

set.seed(42)

# library('ggplot2')
# 
# # --- 1. Set Parameters ---
# 
# # Set a seed for reproducibility
# set.seed(42)
# 
# # Define the number of cells for each region
# n_main <- 2000      # Cells in the general tumor microenvironment
# n_hotspot <- 300    # Cells in the lymphocyte-infiltrated hotspot
# n_stroma <- 500     # Cells in the stroma-dense region
# 
# # Define the four cell types
# cell_types <- c("TC", "Lym", "Stroma", "Others")
# 
# 
# # --- 2. Generate Data for Each Spatial Region ---
# 
# # Region 1: General Tumor Microenvironment (TME)
# # High proportion of TC, moderate Stroma, low Lym and Others.
# dat_main <- data.frame(
#   x = round(runif(n = n_main, min = 1, max = 10000)),
#   y = round(runif(n = n_main, min = 1, max = 10000)),
#   cell_id = sample(
#     cell_types,
#     size = n_main,
#     replace = TRUE,
#     prob = c(0.50, 0.10, 0.30, 0.10) # 50% TC, 10% Lym, 30% Stroma, 10% Others
#   )
# )
# 
# # Region 2: Lymphocyte-rich Hotspot
# # Simulates a strong immune infiltrate with a very high proportion of Lymphocytes.
# dat_hotspot <- data.frame(
#   x = round(runif(n = n_hotspot, min = 1500, max = 4000)),
#   y = round(runif(n = n_hotspot, min = 1500, max = 4000)),
#   cell_id = sample(
#     cell_types,
#     size = n_hotspot,
#     replace = TRUE,
#     prob = c(0.10, 0.70, 0.15, 0.05) # 10% TC, 70% Lym, 15% Stroma, 5% Others
#   )
# )
# 
# # Region 3: Stroma-dense Area
# # Simulates surrounding connective tissue with a high proportion of Stroma.
# dat_stroma <- data.frame(
#   x = round(runif(n = n_stroma, min = 6000, max = 9500)),
#   y = round(runif(n = n_stroma, min = 5000, max = 8500)),
#   cell_id = sample(
#     cell_types,
#     size = n_stroma,
#     replace = TRUE,
#     prob = c(0.15, 0.05, 0.75, 0.05) # 15% TC, 5% Lym, 75% Stroma, 5% Others
#   )
# )
# 
# 
# # --- 3. Combine and Finalize Data ---
# 
# # Combine the data frames from all regions
# dat <- rbind(dat_main, dat_hotspot, dat_stroma)
# 
# # Shuffle the rows to ensure the data isn't ordered by region
# dat <- dat[sample(nrow(dat)), ]
# 
# colnames(dat) <- c("x", "y", "cell_id")
# 
# 
# # --- 4. Visualize the Generated Data ---
# 
# # Create a scatter plot to visualize the spatial distribution of cell types
# cell_plot <- ggplot(dat, aes(x = x, y = y, color = cell_id)) +
#   geom_point(alpha = 0.8, size = 1.5) +
#   coord_fixed() + # Ensure the aspect ratio is 1:1
#   scale_color_manual(values = c(
#     "TC" = "firebrick",
#     "Lym" = "steelblue",
#     "Stroma" = "forestgreen",
#     "Others" = "grey50"
#   )) +
#   labs(
#     title = "Spatially Simulated Cell Data",
#     x = "X Coordinate",
#     y = "Y Coordinate",
#     color = "Cell Type"
#   ) +
#   theme_minimal() +
#   theme(
#     legend.position = "bottom",
#     plot.title = element_text(hjust = 0.5, face = "bold", size = 16)
#   )
# 
# # Print the plot
# print(cell_plot)
# 
# # Display the first few rows of the final data frame
# print(head(dat))
# 
# write.csv(dat, '/Users/lix233/SpatialQPFs_2.0.0/inst/extdata/dat.csv', row.names = F)

dat = read.csv(system.file("extdata", "dat.csv", package = "SpatialQPFs"))


cat("The first 20 cell in the example csv file looks like:")
dat[1:20,]


```


## Visualization of the raw cell data and its spatial density plot

Now we have the generated the data, we want to visualize the cell data, this can be done by calling `Data_Vis()` function. 

* noticed, the "x_var", "y_xar" and "cell_class_var" argument in the functions are the column names that indicting cell position x, y and its cell type, e.g. lymphoctye or tumor-cell, etc. 
* "cell_type" argument indicates which cell type the user wants to visualize.

We look at "Lym" cell type first: 
```{r, warning = FALSE}
path = gsub("dat.csv", "", system.file("extdata", "dat.csv", package = "SpatialQPFs"))
# print(paste0("Your file path is: ", path))
file = "dat.csv"

Data_Vis(path = path, file = file, cell_class_var = "cell_id", x_var = "x", y_var = "y", cell_type = "Lym")  

```

Next, "TC" cell type: 
```{r, warning = FALSE}
Data_Vis(path = path, file = file, cell_class_var = "cell_id", x_var = "x", y_var = "y", cell_type = "TC")  

```


Next, "Stroma" cell type: 
```{r, warning = FALSE}
Data_Vis(path = path, file = file, cell_class_var = "cell_id", x_var = "x", y_var = "y", cell_type = "Stroma")  

```



Next, "Others" cell type: 
```{r, warning = FALSE}
Data_Vis(path = path, file = file, cell_class_var = "cell_id", x_var = "x", y_var = "y", cell_type = "Others")  

```

## Point pattern data analysis

To utilize spatial point process methods, SpatialQPFs provides 3 functions:

* for a single cell type, one can use `Point_pattern_data_uni()`
* for a pair of cell types, one can call `Point_pattern_data_bi()` and `Point_pattern_data_ITLR()`. 
For details of the functions, refer to `help(Point_pattern_data_uni)`, `help(Point_pattern_data_bi)`, and `help(Point_pattern_data_ITLR)`


For "Lym" cell type spatial features:
```{r, warning = FALSE}

ss_point_uni_Lym = Point_pattern_data_uni(path = path, file = file, x_var = "x", y_var = "y", cell_class_var = "cell_id", cell_type = "Lym", scale = 500, myplot = T)

print(ss_point_uni_Lym)
    
```

For "TC" cell type spatial features:
```{r, warning = FALSE}
ss_point_uni_TC = Point_pattern_data_uni(path = path, file = file, x_var = "x", y_var = "y", cell_class_var = "cell_id", cell_type = "TC", scale = 500, myplot = T)

print(ss_point_uni_TC)
    
```

For "Stroma" cell type spatial features:
```{r, warning = FALSE}

ss_point_uni_Lym = Point_pattern_data_uni(path = path, file = file, x_var = "x", y_var = "y", cell_class_var = "cell_id", cell_type = "Stroma", scale = 500, myplot = T)

print(ss_point_uni_Lym)
    
```
For "Others" cell type spatial features:
```{r, warning = FALSE}

ss_point_uni_Lym = Point_pattern_data_uni(path = path, file = file, x_var = "x", y_var = "y", cell_class_var = "cell_id", cell_type = "Others", scale = 500, myplot = T)

print(ss_point_uni_Lym)
    
```



For spatial interaction features between "TC" and "Lym":
```{r, warning = FALSE}
ss_point_bi = Point_pattern_data_bi(path = path, 
                                    file = file, x_var = "x", y_var = "y", cell_class_var = "cell_id", 
                                    from_type = "Lym", 
                                    to_type = "TC", 
                                    scale = 500, myplot = T)

print(ss_point_bi)


ss_point_ITLR = Point_pattern_data_ITLR(path = path, 
                                        file = file, x_var = "x", y_var = "y", cell_class_var = "cell_id", 
                                        from_type = "Lym", 
                                        to_type = "TC", 
                                        micron_per_pixel = 0.5, 
                                        myplot = T)
print(ss_point_ITLR)

```


## Areal data

To utilize spatial lattice process methods, SpatialQPFs provides `Areal_data()` function. For details of the function, refer to `help(Areal_data)`

```{r, warning = FALSE}
ss_lattice_bi = Areal_data(path = path, file = file, 
                           x_var = "x", y_var = "y", cell_class_var = "cell_id", 
                           from_type = "Lym", 
                           to_type = "TC", 
                           scale = 200, 
                           myplot = T)

print(ss_lattice_bi)

```




## Geostatistics data

To utilize geostatistics process methods, SpatialQPFs provides `Geostatistics_data()` function. For details of the function, refer to `help(Geostatistics_data)`

```{r, warning = FALSE}
ss_geostat_bi = Geostatistics_data(path = path, file = file, 
                                   x_var = "x", y_var = "y", cell_class_var = "cell_id", 
                                   from_type = "Lym", 
                                   to_type = "TC", 
                                   scale = 500, 
                                   myplot = T)

print(ss_geostat_bi)


```


## Graph data
```{r, warning = FALSE}

graph_uni_TC = DT_graph_uni_subregion_random(path = path, file = file, 
                                   x_var = "x", y_var = "y", cell_class_var = "cell_id", 
                                   cell_type = "TC", 
                                   scale = 1000, side_length = 2000, num_FOV = 20, set_seed = 42, myplot = T)

graph_uni_Lym = DT_graph_uni_subregion_random(path = path, file = file, 
                                   x_var = "x", y_var = "y", cell_class_var = "cell_id", 
                                   cell_type = "Lym", 
                                   scale = 1000, side_length = 2000, num_FOV = 20, set_seed = 42, myplot = T)


graph_all = DT_graph_cross_subregion_random(path = path, file = file, 
                                   x_var = "x", y_var = "y", cell_class_var = "cell_id", 
                                   cell_type = c("TC", "Lym", "Stroma", "Others"), 
                                   scale = 1000, side_length = 2000, num_FOV = 20, set_seed = 42, myplot = T)


```

## Entropy measurements
### Spatial Entropy using all cell types
```{r, warning=FALSE}
spat_entropy_all = spat_entropy_alltype(path = path, file = file, 
                                        cell_class_var = "cell_id", 
                                        x_var = "x", y_var = "y", scale = 200, 
                                        side_length = 2000, num_FOV = 10, set_seed = 42, myplot = T)
print(spat_entropy_all)
```
### Spatial Entropy using one cell type
```{r, warning=F}
spat_entropy_TC = spat_entropy_onetype(path = path, file = file, 
                                       cell_class_var = "cell_id", 
                                       x_var = "x", y_var = "y", 
                                       cell_type = "TC", 
                                       scale = 200, myplot = T)
  
print(spat_entropy_TC)

spat_entropy_Lym = spat_entropy_onetype(path = path, file = file, 
                                       cell_class_var = "cell_id", 
                                       x_var = "x", y_var = "y", 
                                       cell_type = "Lym", 
                                       scale = 200, myplot = T)
  
print(spat_entropy_Lym)

```

### Local Entropy
```{r, warning=F}

spat_local_entropy = local_entropy(path = path, file = file, 
                                   cell_class_var = "cell_id", 
                                   x_var = "x", y_var = "y", 
                                   scale = 200, myplot = T)
  
spat_local_entropy_TC = spat_local_entropy[dat$cell_id == "TC"]

spat_local_entropy_Lym  = spat_local_entropy[dat$cell_id == "Lym"]
  
# print(spat_local_entropy)  
```


# Aggregating spatial feature outputs for downstream analysis
```{r}
this_res = data.frame(cbind(data.frame("file" = file),
                            data.frame(ss_point_uni_TC) %>% setnames(paste0(names(.),"_TC")), 
                            data.frame(ss_point_uni_Lym) %>% setnames(paste0(names(.),"_Lym")), 
                            data.frame(ss_point_bi), 
                            data.frame(ss_point_ITLR), 
                            data.frame(ss_lattice_bi), 
                            data.frame(ss_geostat_bi),
                            data.frame(graph_uni_TC) %>% setnames(paste0(names(.),"_TC")),
                            data.frame(graph_uni_Lym) %>% setnames(paste0(names(.),"_Lym")),
                            data.frame(graph_all), 
                            data.frame(spat_entropy_all),
                            data.frame(spat_entropy_TC) %>% setnames(paste0(names(.),"_TC")),
                            data.frame(spat_entropy_Lym) %>% setnames(paste0(names(.),"_Lym")),
                            distribution_summary(spat_local_entropy_TC[!is.na(spat_local_entropy_TC)]) %>% setnames(paste0(names(.),"_le_TC")),
                            distribution_summary(spat_local_entropy_Lym[!is.na(spat_local_entropy_Lym)]) %>% setnames(paste0(names(.),"_le_Lym"))
                            ),
                      row.names = NULL)

print(this_res)
```
